---
title: "Statistical inference course project | Part 2"
author: "Mick van Hulst"
date: "30 Jan 2016"
output: pdf_document
---

##1. Load the ToothGrowth data and perform some basic exploratory data analyses
Load data and libraries.
```{r}
library(ggplot2)
library(datasets)

x <- ToothGrowth
```

Convert dose to a factor
```{r}
x$dose <- as.factor(x$dose)
```

Show structure of object 'x'.
```{r}
str(x)
```

Show table
```{r}
table(x$dose, x$supp)
```

A few plots that show the tooth length in comparison with the supplement type and the dose of a certain supplement that is taken.
```{r}
ggplot(data=x, aes(x=dose, y=len, fill=supp)) + geom_bar(stat="identity",) + facet_grid(. ~ supp) + xlab("Dose in miligrams") + ylab("Tooth length") + guides(fill=guide_legend(title="Supplement type"))
```

```{r}
ggplot(aes(x=dose, y=len), data=x) + geom_boxplot(aes(fill=dose)) + xlab("Dose in miligrams") + ylab("Tooth length") + guides(fill=guide_legend(title="Dose"))
```

```{r}
ggplot(aes(x=supp, y=len), data=x) + geom_boxplot(aes(fill=supp)) + xlab("Supplement type") + ylab("Tooth length") + guides(fill=guide_legend(title="Supplement type"))
```

##2. Provide a basic summary of the data
Call the summary function to show a summary.
```{r}
summary(x)
```
##3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering)
Perfm one and two sample t-tests on vector.
```{r}
t.test(len ~ supp, data = x)
```

The null hypothesis cannot be rejected since confindence intervals contain zero and p-value is 0.06. So supplement types seem to have no
impact on tooth growth.

Three groups as per dose level pairs
```{r}
x.doses_0.5_1.0 <- subset (x, dose %in% c(0.5, 1.0)) 
x.doses_0.5_2.0 <- subset (x, dose %in% c(0.5, 2.0)) 
x.doses_1.0_2.0 <- subset (x, dose %in% c(1.0, 2.0)) 
```

Check for the (0.5,1.0) dose levels.
```{r}
t.test(len ~ dose, data = x.doses_0.5_1.0)
```

Check for the (0.5,2.0) dose levels.
```{r}
t.test(len ~ dose, data = x.doses_0.5_2.0)
```

Check for the (1.0,2.0) dose levels.
```{r}
t.test(len ~ dose, data = x.doses_1.0_2.0)
```

The p-value is less than 0.05 and confidence intervals don't contain 0. The average tooth length increases with an increasing dose. The null hypothesis can be rejected.

##4. State your conclusions and the assumptions needed for your conclusions.

**Conclusions**
1. Increasing the dose leads to increased tooth growth.
2. Supplement type seems to have no impact on tooth growth.

**Assumptions**
1. Members of the sample population, i.e. the 60 guinea pigs, are representative of the entire population of guinea pigs. This assumption allows us to generalize the results.

2. The experiment was done with a random assignment of guinea pigs with different dose level categories and supplement type to control for confounders that might affect the outcome.

3. For the t-tests, the variances are assumed to be different for the two groups being compared. This assumption is less stronger than the case in which the variances are assumed to be equal.

